Accelerated Inference for Partially Observed Stochastic Processes with Automatic Differentiation

### 2023-12-30 mop/draft.tex

We first analyze the performance of gradient descent, Newton's method, or any iterative method that multiplies the particle estimate of the gradient with the inverse of a matrix with minimum eigenvalue bounded below, in the case where the log-likelihood surface is $\gamma$-strongly convex. In this case, these methods enjoy a linear convergence guarantee:

@@

Consider a variant of Algorithm \ref{alg:ifad} where one stops if the gradient estimate $||g(\theta_n)|| \leq \sigma \epsilon$. Assume $-\ell$ is $\gamma$-strongly convex, $\gamma I \preceq \nabla_\theta^2 (-\ell) \preceq \Gamma I$. $\sigma = \frac{4 \Gamma}{(1-\beta)}> 4$. Define $G$ as in Assumption \ref{assump:local-bounded-derivative}.

Choose $J$ according to the maximum of the $J$ prescribed in Lemmas \ref{lemma:grad_bound} and $\ref{lemma:hess_bound}$ with failure probability $\delta/2$ each. Let $H(\theta_n)$ be a matrix possibly dependent on $\theta_n$ with minimum eigenvalue always greater than some $c>0$. Choose the learning rate $\eta$ and error tolerance $\epsilon$ such that, where $\beta$ is the Armijo condition hyperparameter,
\begin{equation}
    \eta \leq \frac{c(1-\beta)}{\Gamma}, \;\; \epsilon \leq \frac{c(1-\beta)}{2\Gamma}||g(\theta_n)||.
\end{equation}

@@

The gradient stage of IFAD converges linearly to the MLE if (1) the log-likelihood surface is $\gamma$-strongly convex in a neighborhood of the MLE and (2) the IF2 stage of IFAD successfully reaches a (high-probability) basin of attraction of the MLE. This happens fairly often in practice, for example, when sufficient regularity conditions for local asymptotic normality of the MLE hold.

@@

@@

Even with the increased computational cost of automatic differentiation, Algorithm \ref{alg:ifad} is reasonably fast, as by the cheap gradient principle from \cite{kakade2019provably}, a gradient operation takes no more than 6 times the runtime of the particle filter. As such, unlike methods like the marginal differentiable particle filter in \cite{scibior2021dpf} or the the optimal transport of resampling approach of \cite{corenflos21}, our algorithm has runtime \textbf{linear, and not quadratic}, in the number of particles (while alleviating the $O(N^2)$ variance concern of \cite{poyiadjis11} and \cite{corenflos2021ot} and asymptotic bias concern of \cite{blei2018vsmc} through optimizing $\alpha$ for a bias-variance tradeoff).

>>>>

We first analyze the performance of gradient descent, Newton's method, or any iterative method that multiplies the particle estimate of the gradient with the inverse of a matrix with minimum eigenvalue bounded below, in the case where the log-likelihood surface is $\gamma$-strongly convex, and where the gradient estimate is a particle approximation of the score (for example, IFAD-1 and MOP-1). In this case, these methods enjoy a linear convergence guarantee:

@@

Consider a variant of IFAD-1 (Algorithm \ref{alg:ifad}) where one stops if the gradient estimate $||g(\theta_n)|| \leq \sigma \epsilon$. Assume $-\ell$ is $\gamma$-strongly convex, $\gamma I \preceq \nabla_\theta^2 (-\ell) \preceq \Gamma I$. $\sigma = \frac{4 \Gamma}{(1-\beta)}> 4$. Define $G$ as in Assumption \ref{assump:local-bounded-derivative}.

Choose $J$ according to the maximum of the $J$ prescribed in Lemmas \ref{lemma:grad_bound} and (if using a Hessian estimate) $\ref{lemma:hess_bound}$ with failure probability $\delta/2$ each. Let $H(\theta_n)$ be a matrix possibly dependent on $\theta_n$ with minimum eigenvalue always greater than some $c>0$. Choose the learning rate $\eta$ and error tolerance $\epsilon$ such that, where $\beta$ is the Armijo condition hyperparameter,
\begin{equation}
    \eta \leq \frac{c(1-\beta)}{\Gamma}, \;\; \epsilon \leq \frac{c(1-\beta)}{2\Gamma}||g(\theta_n)||.
\end{equation}

@@

The gradient stage of IFAD-1 converges linearly to the MLE if (1) the log-likelihood surface is $\gamma$-strongly convex in a neighborhood of the MLE and (2) the IF2 stage of IFAD successfully reaches a (high-probability) basin of attraction of the MLE. This happens fairly often in practice, for example, when sufficient regularity conditions for local asymptotic normality of the MLE hold.

@@

While this theoretical guarantee currently only holds for IFAD-1, as it forms a particle approximation of the score as in \cite{poyiadjis11}, one can also extend this for a similar guarantee if the bias for the gradient estimate given by MOP-$\alpha$ for $\alpha < 1$ is small enough. For general $\alpha$, including MOP-0, the analysis then becomes more complicated, as one needs to handle the case of biased gradient descent. However, if future measurements $y_{n+1:N}^*$ do not provide a large amount of information on the identification of the current state $x_n$ given past and current measurements $y_{0:n}^*$, as mentioned in \cite{corenflos21}, the bias may be small enough for it to not present too much of a problem.

@@

Even with the increased computational cost of automatic differentiation, Algorithm \ref{alg:ifad} is reasonably fast, as by the cheap gradient principle from \cite{kakade2019provably}, a gradient operation takes no more than 6 times the runtime of the particle filter (in practice, this means a gradient step takes about 3.75x as much time as an IF2 iteration). As such, unlike methods like the marginal differentiable particle filter in \cite{scibior2021dpf} or the the optimal transport of resampling approach of \cite{corenflos21}, our algorithm has runtime \textbf{linear, and not quadratic}, in the number of particles (while alleviating the $O(N^2)$ variance concern of \cite{poyiadjis11} and \cite{corenflos2021ot} and asymptotic bias concern of \cite{blei2018vsmc} through optimizing $\alpha$ for a bias-variance tradeoff).
